---
title: "ps3_MultipleDecrement_b"
name: "Andrea Cristina Ruiz"
date: "2025-09-29"
format: pdf
editor: visual
---

## \-\--

```{r}
#settup 
#install.packages("rmarkdown") 
#install.packages("tinytex") 
#install.packages("miktex") 
#install.packages("Rtools") #something about it not available for my version of R
#install.packages("pdflatex") 
#install.packages("quarto")
#library(HMDHFDplus)
library(dplyr)
library(ggplot2)
library(tidyverse) 
library(tinytex)
library(miktex)
library(pdflatex)
library(quatro)
```

---
## Second-level header {#introduction .important .large}
Folded: Downloading the data
---

Downloading the data and keeping only 2005 life table data

```{r}
##"Step 1: Obtain the 2005 single year age-specific death probabilities from the Human Mortality Databas"
#death probability is q_x

####Question 1 get life table values directly.
  # Set your HMD credentials. attaches values to names with my creds. will give that info to the webpage when i request entry

  #next, pull the data. Idk why, but it seems important to run the library before
    library(HMDHFDplus)
      #item: "blt" is building life table, "per" is period life table, "1x1" ys the year increments. 
    
    #for USA ALL in 2005 tables. Combines Both sexes
    usa_data_pull <- readHMDweb(CNTRY = "USA", 
                                 item = "bltper_1x1", # Period life tables by single year
                           username = username,
                           password = password)
                                    
      #for USA males in 2005 tables change item to 
                 #for males item = "mltper_1x1", 
                  #for females item = "fltper_1x1",  
   
#filter to keep only 2005. can use base or dplyr. try using dplyr. "filter() function is used to subset a data frame, retaining all rows that satisfy your conditions. To be retain"
    
    usa_2005 <- usa_data_pull %>%
               filter(Year == 2005) 
    
               View(usa_2005)  #check. I see 111 observations all 2005. Great!
               
   #clean up. remove things i dont need 
     rm(usa_data_pull, password, username, usa_data_pull)
```

Question 1a) What proportion of Wisconsinites who live to age 16 will live to age 31 without experiencing a motor vehicle crash?\
Answer: 17.92693

Process: A proportion requires a denominator and a numerator. In this problem, the **denominator**, $l_0$ will be 85,000, total number of people alive at age 16

The **numerator** will be the number of people who **live** to age 31 without a crash. Sum of people who don't die/ people at age 16. Ingredients:

-   Not dying from a car crash in period is estimated as the total number of people minus people dying from cause i (crash) at age n. $l_n -l^i_{n}*p^i_n$ Surviving one period = not dying in one period.

-   Sum this up for every year between 16 and 31. $\Sigma _{n=16}^{31} [l_n - (l^i_{n}*p^i_n)]$

    -   The probability that someone crashes (i) at any age is $q^i_x =0.062 - 0.000053*(\textrm{age} ^2)$. Applying the formula at age 16 that is $0.062 - 0.000053*(16^2)$,

    -   Deaths at each interval are $d_x = l^i_{n}*p^i_n$

    -   Number of people not dying from crashes is $l_n-l^i_n$ \### NOT SO SURE ABOUT THIS CALCULATION. Confirm.

```{r}
#first make a new variable for the probability of people dying from crashes in each age . 
            #didnt work because it didnt call a data frame. QUESTION- how would I update this equation to work better? 
      #prob_crash_die = 0.062 - 0.000053 * (Age^2)

#second try. mutate the data frame 
usa_2005 <- usa_2005 %>%
     mutate(prob_crash_die = 0.062 - 0.000053 * (Age^2))

#second, calculate how many people die from crashes at each age interval. l_n *prob dying from crash.  
usa_2005 <- usa_2005 %>%
     mutate(crash_deaths = prob_crash_die*lx)

#Question. should it be ppl from the year before. 15 yr old alive * 16 yr old prob dying??

#third find out who does not die from crashes per year. Aka- who lives. 
usa_2005 <- usa_2005 %>%
     mutate(ln_minus_crshdeath = lx - crash_deaths)
      #add new var with better name 
     usa_2005 <- usa_2005 %>%
           mutate(didnt_die = ln_minus_crshdeath)

#sum up people who didnt die between 16 and 31. 
     #had issues with this so broke it down to step by step
     test_ages_14 <- usa_2005$Age >= 0 & usa_2005$Age <= 4
      head(test_ages) #it worked 
      
      sum_didntDie_16_31 <- sum(usa_2005[usa_2005$Age >= 16 & usa_2005$Age <= 31, "didnt_die"])
      sum_didntDie_16_31  #it worked 1523789
      
#finally, the equation. 17.92693
    answ_1a <-  sum_didntDie_16_31/85000
      answ_1a
  
```

B\) Among those who live to age 25 accident-free, what is the probability of experiencing an accident by age 31? Answer: 50.56%

QUESTION, PROB CRASH IS SUM OF PROBS WITHIN SUB POP. NO DENOM

Numerator: Probability of experiencing crash between 25 and 31$(p^i_{16} +p^i_{17}….p^i_{31})$

```{r}
sum_probdie_16_31 <- sum(usa_2005[usa_2005$Age >= 16 & usa_2005$Age <= 31, "prob_crash_die"])
      sum_probdie_16_31

```

C. Among those who survive to age 16, what is the probability of dying without experiencing an accident by age 31?

Approach 1: (probability of dying between 16 and 31)- (probability of dying in crash between 16 and 31)

pro

(dying/total)-(crash dying/total)

```{r}
 
#dying between 16 and 31 
sum_died_16_31 <- sum(usa_2005[usa_2005$Age >= 16 & usa_2005$Age <= 31, "dx"])

#crash dying
sum_probdie_16_31 <- sum(usa_2005[usa_2005$Age >= 16 & usa_2005$Age <= 31, "prob_crash_die"])
```

D. If the experience of accidents and the probability of dying are process-dependent, is your estimate for C an overestimate or an underestimate of the true probability?

If a experiencing a crash and the probability of dying is positively correlated, then the estimate is an underestimate. If the correlation is negative, 1c is an over estimate.

E. Extra Credit: Push your code to GitHub and share the link with someone from class. Answer here the name of the person(s) to whom you shared the link.

```{r}
#first time pushing code through quatro. ACR NOTE UPDATE THIS
cd /path/to/your/project
git init
# Or add all files
git add .

# Make your first commit
git commit -m "Add initial .qmd file"

# Add remote repository (replace with your GitHub repo URL)
git remote add origin https://github.com/yourusername/your-repo-name.git

# Push to GitHub
git push -u origin main

#ALTERNATIVE QUIT.
git add . && git commit -m "Update qmd file" && git push origin main
# Replace main with master if your default branch is named master, and replace the repository URL with your actual GitHub repository URL.
```

The `echo: false` option disables the printing of code (only output is displayed).

