---
title: "ps3_MultipleDecrement_b"
name: "Andrea Cristina Ruiz"
date: "2025-09-29"
format: html
---

## ---

Set up install packages

```{r}
#settup 

library(rmarkdown)
library(HMDHFDplus)
library(dplyr)
library(ggplot2)
library(tidyverse) 
#library(TinyTex)
#library(miktex)
#library(pdflatex)
library(quarto)
library(tidyverse)

options(repos = c(CRAN = "https://cloud.r-project.org/"))
```

Downloading the data and keeping only 2005 life table data

```{r}
##"Step 1: Obtain the 2005 single year age-specific death probabilities from the Human Mortality Databas"
#death probability is q_x

####Question 1 get life table values directly.
  # Set your HMD credentials. attaches values to names with my creds. will give that info to the webpage when i request entry
   # username <-  DELETED
   # password <-  DELETED
    
  #next, pull the data. Idk why, but it seems important to run the library before
    #library(HMDHFDplus)
      #item: "blt" is building life table, "per" is period life table, "1x1" ys the year increments. 
 
    #for USA ALL in 2005 tables. Combines Both sexes
   # usa_data_pull <- readHMDweb(CNTRY = "USA", 
                    #             item = "bltper_1x1",
                     #      username = username,
                      #     password = password)
                                    
      #for USA males in 2005 tables change item to 
                 #for males item = "mltper_1x1", 
                  #for females item = "fltper_1x1",  
   
#filter to keep only 2005. can use base or dplyr. try using dplyr. "filter() function is used to subset a data frame, retaining all rows that satisfy your conditions. To be retain"
    
   # usa_2005 <- usa_data_pull%>% 
      #        filter(Year == 2005)
    
              # View(usa_2005)  #check. I see 111 observations all 2005. Great!
               
   #clean up. remove things i dont need 
   #  rm(usa_data_pull, password, username, usa_data_pull)
     
     #worke great all last week but didnt work on 9.30. used excel. 
usa_2005 <- read.csv("~/PhD Courses/2025 F Courses/Demography/HW/usa_2005_test.csv", sep="")

 head (usa_2005)          
      saveRDS(usa_2005, file = "usa_2005.rds")

    # Load it back later
    usa_2005 <- readRDS("usa_2005.rds")
    class(usa_2005$Age)
    usa_2005[] <- lapply(usa_2005, as.numeric) #issue for 100+ but that is not relevant for today
    usa_2005$Age[is.na(usa_2005$Age)] <- 110 #remove NAs
```

Question 1a) What proportion of Wisconsinites who live to age 16 will live to age 31 without experiencing a motor vehicle crash?\
Answer:

```         
0.5956642
```

Process: A proportion requires a denominator and a numerator. In this problem, the **denominator**, "$l_0$" will be 85,000, total number of people alive at age 16

The **numerator** will be the number of people who **live** to age 31 without a crash, or $l_{16}-l_{31}$ . To find $l_{31}$ I do the following:

-   Estimate crash risk by age to find $q^i_x$ which is probability of a crash and probability of dying (sum)

```{r}
#first make a new variable for the probability of people dying from crashes in each age . 
            #didnt work because it didnt call a data frame. QUESTION- how would I update this equation to work better? 
      #prob_crash_die = 0.062 - 0.000053 * (Age^2)

#second try #mutate the data frame 
usa_2005 <- usa_2005 %>%
     mutate(qx_crash = 0.062 - 0.000053 * (Age^2))


#calculate probability of crash + probabilit of dying. q+xdie + q_xcrash
  usa_2005 <- usa_2005 %>%
         mutate(qx_crash_n_die = qx+qx_crash)


#make a wisconsin lx starting at age 16.
     library(dplyr)
      usa_2005 <- usa_2005 %>%
        mutate(lx_wi = ifelse(Age == 16, 85000, NA))
      
      #probability of dying is lx0*(1-qx_crash_n_die. Start by calculating age 17 
        # Calculate Age 17
        usa_2005$lx_wi[usa_2005$Age == 17] <- 
              usa_2005$lx_wi[usa_2005$Age == 16] * 
              (1 - usa_2005$qx_crash_n_die[usa_2005$Age == 16])
        #extend to all ages 
          usa_2005 <- usa_2005 %>%   #first arragnge in order
            arrange(Age)
        #next- calculate with loop .

          for(age in 18:110) {
              usa_2005$lx_wi[usa_2005$Age == age] <- 
                usa_2005$lx_wi[usa_2005$Age == (age - 1)] * 
                (1 - usa_2005$qx_crash_n_die[usa_2005$Age == (age - 1)])
            }  #omg it worked yay

#now that I have lx. I can find the numerator.         
      lxwi16_lxwi31 <- usa_2005$lx_wi[usa_2005$Age == 16] %>%-usa_2005$lx_wi[usa_2005$Age == 31]
      
#divide Numerator by denominator 
     alive31_crash<- (lxwi16_lxwi31 /85000)
      print (alive31_crash)

    #alive without crash 
      alive31_nocrash <- (1-alive31_crash)
print(alive31_nocrash)

```

B\) Among those who live to age 25 accident-free, what is the probability of experiencing an accident by age 31?

Denominator is number of Wi people

Numerator: Probability of experiencing crash between 25 and 31 is the sum of people experiencing $(p^i_{25}*l^{crash}_{25} +p^i_{26}*l^{crash}_{26}….+ p^i_{31}*l^{crash}_{31})$

```{r}

#already calculated probability of crash qx_crash. now estimate dx_crash, number of people* prob of crash. Then sum. 

usa_2005 <- usa_2005 %>%
  mutate(dx_crash = lx_wi * qx_crash)

#note, dont include 31
dxcrash_25_30 <- sum(usa_2005[usa_2005$Age >= 25 & usa_2005$Age <= 30, "dx_crash"])     
print(dxcrash_25_30)
```

Denominator: Number of people in WI alive at 25

```{r}

deonm1b <- usa_2005$lx_wi[usa_2005$Age == 25]

ans_1b <- dxcrash_25_30/deonm1b

print(ans_1b)
```

C. Among those who survive to age 16, what is the probability of dying without experiencing an accident by age 31?

Numerator: (probability of dying between 16 and 31)- (probability of dying in crash between 16 and 31)

Denominator: $l_{16}$

```{r}
 #need dx_wi 
  # made a mistake first time. removing var 
  #usa_2005$dx_wi <- NULL
  
usa_2005 <- usa_2005 %>%
  mutate(dx_wi = (qx*lx_wi)) #prob dying in usa * # ppl in Wi

    #did it work? It did!
    usa_2005 %>%
      filter(Age >= 16, Age <= 30) %>%
       
 
#find died and accident 
    usa_2005 <- usa_2005 %>%
          mutate(dx_diencrsh = (dx_wi+dx_crash))
    
#dying and crash between 16 and 31 
sum_diencrsh_16_30 <- sum(usa_2005[usa_2005$Age >= 16 & usa_2005$Age < 31, "dx_diencrsh"])

#crash experience, non death
sum_crash_16_31 <- sum(usa_2005[usa_2005$Age >= 16 & usa_2005$Age < 31, "dx_crash"])

#next find numerator. 
num_1c <- sum_diencrsh_16_30 -sum_crash_16_31 

Ans1c <- (num_1c/85000)

print(num_1c)
print(Ans1c)
```

D. If the experience of accidents and the probability of dying are process-dependent, is your estimate for C an overestimate or an underestimate of the true probability?

If a experiencing a crash and the probability of dying is positively correlated, then the estimate is an underestimate. If the correlation is negative, 1c is an over estimate.

E. Extra Credit: Push your code to GitHub and share the link with someone from class. Answer here the name of the person(s) to whom you shared the link.

```{r}
#first time pushing code through quatro.  
getwd()

#"C:/Users/Andrea Cristina/OneDrive - UW-Madison/Documents/PhD Courses/2025 F Courses/Demography/HW"


```

The `echo: True`option disables the printing of code (only output is displayed).
